The Impact of Alternation

### 2017-05-15 draft.tex
ok

\author{Radu Iosif and Xiao Xu \\
Verimag/CNRS/Universit\'e de Grenoble Alpes\\
Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr}

@@

\paragraph{\em Remark}
While computing the concrete post-image $\Post{\A}$ relies on an
expensive quantifier elimination, it is common belief that this should
hold also for $\AbsPost{\A}$. However, chosing the conjoined
predicates that form $\AbsPost{\A}$ requires only a quantifier-free
satisfiability query: $\Post{\A}(\phi,a) \models \pi$ iff $\exists
\overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}],a)
\models \pi$ iff $\Delta(\phi[\overline{\vec{x}}/\vec{x}],a) \wedge
\neg\pi$ is unsatisfiable. \qed

@@

\paragraph{\em Remark} 
We deviate w.l.o.g. from the standard notion of \emph{generalized
  interpolants} \cite{mcmillan06}, which requires $\fv{}{I_i}
\subseteq \fv{}{\phi_i} \cap \fv{}{\phi_{i+1}}$, for each $i \in
          [1,n-1]$, and assume further that the variables have been
          renamed by removing the time stamps. \qed

@@

\begin{figure}[htb]
\begin{tabular}{cc}
\includegraphics[scale=0.52]{figures/PA1.pdf} & \includegraphics[scale=0.52]{figures/PA2.pdf} \\
\footnotesize{(a)} & \footnotesize{(b)} \\
\\
\includegraphics[scale=0.52]{figures/PA3.pdf} & \includegraphics[scale=0.52]{figures/PA4.pdf} \\
\footnotesize{(c)} & \footnotesize{(d)}
\end{tabular}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} using Algorithm \ref{alg:predabs}}
\label{fig:pa}
\end{figure}

@@

@@

%% The ART built by Algorithm \ref{alg:impact} corresponds to an
%% unfolding of the input automaton by enumerating the input event
%% sequences in $\Sigma^*$. Each time a new node is created, its label is
%% initially $\top$. This label is subject to subsequent strenghtening,
%% when a spurious counterexample is encountered. To speedup the
%% computation of interpolants for spurious counterexamples, we use two
%% additional labeling functions for bookkeeping. Formally, the ART
%% structure used by Algorithm \ref{alg:impact} is defined as $\Art =
%% \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where $N,E,\Lambda$ and
%% $\lhd$ are defined as for Algorithm \ref{alg:predabs} and,
%% moreover: \begin{compactitem}
%% \item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
%%   sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$, 
%% %
%% \item $T : E \rightarrow \bigcup_{i=0}^\infty
%%   \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges
%%   with time-stamped formulae, and
%% %
%% \item for each edge $e = (n,a,m) \in E$, if $n$ lies at
%%   depth\footnote{We consider that the root node has zero depth.}  $k$
%%   in the tree $(N,E)$, then we have:
%% \end{compactitem}
%%   \[\begin{array}{rcl}
%%   T(e) & = & \Theta(n,a,k) \\
%%   R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{, where} \\
%%   \Theta(n,a,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
%%   \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}])
%%   \end{array}\]
%% For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is
%% $(n,a,m)$. This additional bookkeeping is used as follows. Given a
%% sequence $u = a_1\ldots a_k \in \Sigma^*$ that labels the path from
%% the root to a node $n$ in $\Art$, the conjunction:
%% \[\mu(n) \equiv \Lambda(\rootNode) \wedge T(\rootNode,a_1,n_1) \wedge \ldots
%% T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q_k
%% \rightarrow \bot)\] is equivalent to the interpolation problem
%% $\pi(u)$ and can be directly computed by a traversal of the path from
%% $\rootNode$ to $n$.

@@

\begin{figure}[t!]
\begin{tabular}{cc}
\includegraphics[scale=0.42]{figures/IM1.pdf} & \includegraphics[scale=0.42]{figures/IM2.pdf} \\
\footnotesize{(a)} & \footnotesize{(b)} \\
\\
\includegraphics[scale=0.42]{figures/IM3.pdf} & \includegraphics[scale=0.42]{figures/IM4.pdf} \\
\footnotesize{(c)} & \footnotesize{(d)}
\end{tabular}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} using Algorithm \ref{alg:impact}}
\label{fig:im}
\end{figure}

\paragraph{\em Example}
We show the execution of Algorithm \ref{alg:impact} on the automaton
from Figure \ref{fig:ex}.  Initially, the procedure fires the sequence
$a$, whose endpoint is labeled with $\top$, in Figure \ref{fig:im}
(a). Since this node is uncovered, we check the spuriousness of the
counterexample $a$ and refine the label of the node to $q_1$. Since
the node is still uncovered, two successors, labeled with $\top$ are
computed, corresponding to the sequences $aa$ and $ab$, in Figure
\ref{fig:im} (b). The spuriousness check for $aa$ yields the
interpolant $\tuple{q_0,x\leq0\wedge q_2 \wedge y\geq0}$ which
strenghtens the label of the endpoint of $a$ from $q_1$ to $q_1\wedge
x\leq0\wedge q_2\wedge y\geq0$. The sequence $ab$ is also found to be
spurious, which changes the label of its endpoint from $\top$ to
$\bot$, and also covers it (depicted with a dashed edge). Since the
endpoint of $aa$ is not covered, it is expanded to $aaa$ and $aab$, in
Figure \ref{fig:im} (c). Both sequences $aaa$ and $aab$ are found to
be spurious, and the enpoint of $aab$, whose label has changed from
$\top$ to $\bot$, is now covered. In the process, the label of $aa$
has also changed from $q_1$ to $q_1 \wedge y>x-1 \wedge q_2$, due to
the strenghtening with the interpolant from $aab$.  Finally, the only
uncovered node $aaa$ is expanded to $aaaa$ and $aaab$, both found to be
spurious, in Figure \ref{fig:im} (d). The refinement of $aaab$ causes
the label of $aaa$ to change from $q_1$ to $q_1 \wedge y>x-1 \wedge
q_2$ and this node is now covered by $aa$. Since its successors are
also covered, there are no uncovered nodes and the procedure returns
$\true$. \qed

The following lemma states that the labeling of the nodes in an ART
built by Algorithm \ref{alg:impact} is consistent with the semantics
of the transition function of $\A$.

\begin{lemma}\label{lemma:well-labeled}
  Given an ART $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ built
  by Algorithm \ref{alg:impact}, $\Post{\A}(\Lambda(n),a) \models
  \Lambda(m)$, for all $(n,a,m) \in E$.
\end{lemma}

@@

@@

\subsection{Proof of Lemma \ref{lemma:well-labeled}}

\proof{ We distinguish two cases. First, if $(n,a,m)$ occurs on a path
  in $\Art$ that has never been refined, then $\Lambda(m)=\top$ and
  the entailment holds trivially. Otherwise, let $\Omega$ be the set
  of paths $\omega = (n_0,a_1,n_1), \ldots, (n_{k-1},a_k,n_k)$, where
  $n_0 = \rootNode$ and $(n,a,m) = (n_{i-1},a_i,n_i)$, for some $i \in
  [1,k]$ and, moreover, $a_1\ldots a_k$ was found, at some point, to
  be a spurious counterexample. Let
  $\langle\top,I^\omega_0,\ldots,I^\omega_k,\bot\rangle$ be an
  interpolant for $\Phi(a_1 \ldots a_k) \equiv \Lambda(\rootNode)
  \wedge \bigwedge_{i=1}^k \theta_i \wedge \bigwedge_{q \in R(n_k)}
  (q_k \rightarrow \bot)$, such that $I^\omega_i \in
  \Form^+(Q,\vec{x})$, for all $i \in [0,k]$. According to Proposition
  \ref{prop:positive-interpol}, it is possible to build such an
  interpolant, when $\Phi(a_1 \ldots a_k)$ is unsatisfiable. By
  Proposition \ref{prop:asterisk}, we obtain
  $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \iffeq \exists Q_{i-1} ~.~
  I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge \theta_i$
  and, since $I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge
  \theta_i \models I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$, we obtain
  that $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \models
  I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$. Since $\Lambda(n_{i-1}) =
  \bigwedge_{\omega \in \Omega} I^\omega_{i-1}$ and $\Lambda(n_{i}) =
  \bigwedge_{\omega \in \Omega} I^\omega_{i}$, we obtain
  $\Post{\A}(\Lambda(n_{i-1}),a_i) \models \Lambda(n_i)$. \qed}

@@

\proof{ We prove first that, each time Algorithm \ref{alg:impact}
  reaches the line \ref{ln:impact-while}, we have:
  \begin{equation}\label{inv:impact}
  W = \set{n \mid n \text{ uncovered, } \exists a \in \Sigma~ \forall
    s \in N ~.~ (n,a,s) \not\in E}
  \end{equation}
  Initially, $W = \set{\rootNode}$ and $E=\lhd=\emptyset$, thus
  (\ref{inv:impact}) holds trivially. Suppose that (\ref{inv:impact})
  holds at when reaching line \ref{ln:impact-while} and some node $n$
  was removed from $W$ and inserted into $N$. We distinguish two
  cases, either: \begin{itemize}
  \item $n$ is covered, in which case $W$ becomes $W \setminus
    \set{n}$ and (\ref{inv:impact}) holds, or
  %
  \item $n$ is not covered, in which case $W$ becomes $(W \setminus
    \set{n}) \cup S$, where $S = \set{s \not\in N \mid (n,a,s) \in E,
      a \in \Sigma}$ is the set of fresh successors of $n$. But then
    no node $s \in S$ is covered and has successors in $E$, thus
    (\ref{inv:impact}) holds.
  \end{itemize}
  Then the condition (\ref{inv:impact}) holds next time line
  \ref{ln:impact-while} is reached, thus it is invariant.

>>>>

\author{Radu Iosif and Xiao Xu \\
CNRS, Verimag, Universit\'e de Grenoble Alpes \\
Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr}

@@

%% \paragraph{\em Remark}
%% While computing the concrete post-image $\Post{\A}$ relies on an
%% expensive quantifier elimination, it is common belief that this should
%% hold also for $\AbsPost{\A}$. However, chosing the conjoined
%% predicates that form $\AbsPost{\A}$ requires only a quantifier-free
%% satisfiability query: $\Post{\A}(\phi,a) \models \pi$ iff $\exists
%% \overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}],a)
%% \models \pi$ iff $\Delta(\phi[\overline{\vec{x}}/\vec{x}],a) \wedge
%% \neg\pi$ is unsatisfiable. \qed

@@

%% \paragraph{\em Remark} 
%% We deviate w.l.o.g. from the standard notion of \emph{generalized
%%   interpolants} \cite{mcmillan06}, which requires $\fv{}{I_i}
%% \subseteq \fv{}{\phi_i} \cap \fv{}{\phi_{i+1}}$, for each $i \in
%%           [1,n-1]$, and assume further that the variables have been
%%           renamed by removing the time stamps. \qed

@@

@@

\begin{figure}[t!]
\begin{tabular}{cc}
\includegraphics[scale=0.5]{figures/PA1.pdf} & \includegraphics[scale=0.5]{figures/PA2.pdf} \\[-2mm]
\tiny{(a)} & \tiny{(b)} \\
\\
\includegraphics[scale=0.5]{figures/PA3.pdf} & \includegraphics[scale=0.5]{figures/PA4.pdf} \\[-2mm]
\tiny{(c)} & \tiny{(d)}
\end{tabular}
\vspace*{-\baselineskip}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} by Algorithm \ref{alg:predabs}}
\label{fig:pa}
\end{figure}

@@

@@

\begin{figure}[htb]
\vspace*{-\baselineskip}
\begin{tabular}{c}
\includegraphics[scale=0.65]{figures/IM1.pdf} \\[-2mm]
\tiny{(a)} \\
\includegraphics[scale=0.65]{figures/IM2.pdf} \\[-2mm]
\tiny{(b)} \\
\includegraphics[scale=0.65]{figures/IM3.pdf} \\[-2mm]
\tiny{(c)}
\end{tabular}
\vspace*{-0.5\baselineskip}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
  by Algorithm \ref{alg:impact} (1/2)}
\label{fig:im1}
\end{figure}

\begin{figure}[htb]
\centerline{\includegraphics[scale=0.65]{figures/IM4.pdf}}
\vspace*{-\baselineskip}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
  by Algorithm \ref{alg:impact} (2/2)}
\label{fig:im2}
\end{figure}

\paragraph{\em Example}
We show the execution of Algorithm \ref{alg:impact} on the automaton
from Figure \ref{fig:ex}.  Initially, the procedure fires the sequence
$a$, whose endpoint is labeled with $\top$, in Figure \ref{fig:im1}
(a). Since this node is uncovered, we check the spuriousness of the
counterexample $a$ and refine the label of the node to $q_1$. Since
the node is still uncovered, two successors, labeled with $\top$ are
computed, corresponding to the sequences $aa$ and $ab$, in Figure
\ref{fig:im1} (b). The spuriousness check for $aa$ yields the
interpolant $\tuple{q_0,x\leq0\wedge q_2 \wedge y\geq0}$ which
strenghtens the label of the endpoint of $a$ from $q_1$ to $q_1\wedge
x\leq0\wedge q_2\wedge y\geq0$. The sequence $ab$ is also found to be
spurious, which changes the label of its endpoint from $\top$ to
$\bot$, and also covers it (depicted with a dashed edge). Since the
endpoint of $aa$ is not covered, it is expanded to $aaa$ and $aab$, in
Figure \ref{fig:im1} (c). Both sequences $aaa$ and $aab$ are found to
be spurious, and the enpoint of $aab$, whose label has changed from
$\top$ to $\bot$, is now covered. In the process, the label of $aa$
has also changed from $q_1$ to $q_1 \wedge y>x-1 \wedge q_2$, due to
the strenghtening with the interpolant from $aab$.  Finally, the only
uncovered node $aaa$ is expanded to $aaaa$ and $aaab$, both found to be
spurious, in Figure \ref{fig:im2}. The refinement of $aaab$ causes
the label of $aaa$ to change from $q_1$ to $q_1 \wedge y>x-1 \wedge
q_2$ and this node is now covered by $aa$. Since its successors are
also covered, there are no uncovered nodes and the procedure returns
$\true$. \qed

@@

xxx

@@

@@

\begin{lemma}\label{lemma:well-labeled}
  Given an ART $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ built
  by Algorithm \ref{alg:impact}, $\Post{\A}(\Lambda(n),a) \models
  \Lambda(m)$, for all $(n,a,m) \in E$.
\end{lemma}
\proof{ We distinguish two cases. First, if $(n,a,m)$ occurs on a path
  in $\Art$ that has never been refined, then $\Lambda(m)=\top$ and
  the entailment holds trivially. Otherwise, let $\Omega$ be the set
  of paths $\omega = (n_0,a_1,n_1), \ldots, (n_{k-1},a_k,n_k)$, where
  $n_0 = \rootNode$ and $(n,a,m) = (n_{i-1},a_i,n_i)$, for some $i \in
  [1,k]$ and, moreover, $a_1\ldots a_k$ was found, at some point, to
  be a spurious counterexample. Let
  $\langle\top,I^\omega_0,\ldots,I^\omega_k,\bot\rangle$ be an
  interpolant for $\Phi(a_1 \ldots a_k) \equiv \Lambda(\rootNode)
  \wedge \bigwedge_{i=1}^k \theta_i \wedge \bigwedge_{q \in R(n_k)}
  (q_k \rightarrow \bot)$, such that $I^\omega_i \in
  \Form^+(Q,\vec{x})$, for all $i \in [0,k]$. According to Proposition
  \ref{prop:positive-interpol}, it is possible to build such an
  interpolant, when $\Phi(a_1 \ldots a_k)$ is unsatisfiable. By
  Proposition \ref{prop:asterisk}, we obtain
  $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \iffeq \exists Q_{i-1} ~.~
  I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge \theta_i$
  and, since $I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge
  \theta_i \models I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$, we obtain
  that $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \models
  I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$. Since $\Lambda(n_{i-1}) =
  \bigwedge_{\omega \in \Omega} I^\omega_{i-1}$ and $\Lambda(n_{i}) =
  \bigwedge_{\omega \in \Omega} I^\omega_{i}$, we obtain
  $\Post{\A}(\Lambda(n_{i-1}),a_i) \models \Lambda(n_i)$. \qed}

\vspace*{\baselineskip}
\proof{ We prove first that, each time Algorithm \ref{alg:impact}
  reaches the line \ref{ln:impact-while}, we have:
  \begin{equation}\label{inv:impact}
  W = \set{n \mid n \text{ uncovered, } \exists a \in \Sigma~ \forall
    s \in N ~.~ (n,a,s) \not\in E}
  \end{equation}
  Initially, $W = \set{\rootNode}$ and $E=\lhd=\emptyset$, thus
  (\ref{inv:impact}) holds trivially. Suppose that (\ref{inv:impact})
  holds at when reaching line \ref{ln:impact-while} and some node $n$
  was removed from $W$ and inserted into $N$. We distinguish two
  cases, either: \begin{itemize}
  \item $n$ is covered, in which case $W$ becomes $W \setminus
    \set{n}$ and (\ref{inv:impact}) holds, or
  %
  \item $n$ is not covered, in which case $W$ becomes $(W \setminus
    \set{n}) \cup S$, where $S = \set{s \not\in N \mid (n,a,s) \in E,
      a \in \Sigma}$ is the set of fresh successors of $n$. But then
    no node $s \in S$ is covered and has successors in $E$, thus
    (\ref{inv:impact}) holds.
  \end{itemize}
  Then the condition (\ref{inv:impact}) holds next time line
  \ref{ln:impact-while} is reached, thus it is invariant.
